\section{Key Considerations} \label{sec:key-considerations}

\begin{enumerate} \itshape
\item Using the SDK requires programming in CSL, a C-like language designed specifically for the problem of massively parallel programming on the CS-2.
What is your experience with HPC programming paradigms and languages, such as MPI, OpenMP, CUDA, OpenCL, etc.?
\end{enumerate}

Dr. Moreno extensive experience working with HPC technologies, most notably developing asynchronous, best-effort computing approaches using MPI and C++.
He has also received training in GPU programming (e.g., OpenCL, CUDA).

Dr. Dolson and Dr. Zaman are both experienced systems-level programmers.

\begin{enumerate}[resume] \itshape
\item What are the underlying computational algorithms you're interested in exploring?
What existing software packages or libraries use these algorithms?
\end{enumerate}

The core algorithm is an island-style GA.
i.e., Empirical (cite), MABE (cite), and others.
These frameworks serial processing on traditional HPC hardware.

\begin{enumerate}[resume] \itshape
\item How is this problem bottlenecked on current hardware?
Is the problem more bottlenecked by memory bandwidth, or communication costs associated with scaling up across distributed compute nodes?
\end{enumerate}

Memory bandwidth and synchronization costs to exchange population members.
Compute bound memory latency.

\begin{enumerate}[resume] \itshape
\item What range of problem sizes are you interested in addressing?
For example, how much memory does your problem use?
How does memory usage or program run-time scale with problem size?
\end{enumerate}

We are interested in developing population-scale, cross-scale simulations with hundreds of millions of agents.
The memory usage of the program and run-time scale linearly with problem size, either defined as evolutionary duration or population size.
Precise memory usage depends largely on the agent genomes used.
Including instrumentation, a meaningful simulation size could feasibly be achieved using on the order of 4 words of memory per agent.

\begin{enumerate}[resume] \itshape
\item What portion of your algorithm do you plan to port to the CS-2?
Why are you interested in exploring this part of your algorithm?
\end{enumerate}

We plan to port the core evolutionary generational loop to the CS-2 to as an island model-style genetic algorithm.
The stages are 1) migration, 2) selection, 3) replication with mutation.
We are interested in porting it because it is the most coputationally intensive to scale and because it is embarassingly parallel.

Postprocessing to analyze evolved traits, and reconstruct evolutionary history will take place off-device.

\begin{enumerate}[resume] \itshape
\item The CS-2 offers native half and single precision data types.
What precision does your algorithm or use case need?
\end{enumerate}

Due to the agent-based modeling, this is a primarily discrete domain.
However, floating point types may be used to represent traits.
Becuase they are only used for comparisons between agents, single or half precision suffice.

\begin{enumerate}[resume] \itshape
\item The CS-2 is a network-attached accelerator. At a high level, the CSL programming model is similar to that of CUDA, in which data is moved between host CPU nodes and the device (CS-2) on which computational kernels are launched.
How often will data need to be moved between the wafer and the worker nodes?
\end{enumerate}

For preliminary versions of the project, we are planning to only collect data at the end of execution the through the Cerebras memcpy infrastructure.
As the model is self-contained, no data except for PE code will need to be moved onto the device.

In the future, we are interested in exploring samples throughout the evolutionary process to serve as a kind of ``fossil record.''

\begin{enumerate}[resume] \itshape
\item Describe your general plan to map your problem onto 850,000 cores. In answering this question, it might be helpful to recall some details of the CS-2 architecture.
The 850,000 cores are laid out in a mesh, with each core connected on fabric to its four nearest neighbors on the East, South, North, and West.
Memory is distributed among the cores, with each core having 48 KB of local memory.
64-bit local memory reads and writes take roughly a cycle, as does sending and receiving 32-bit messages between the four neighboring cores.
\end{enumerate}

We plan to exploit an island model genetic algorithm where each PE is an island and genomes migrate between adjacent islands.
As natural systems are typically asynchronous in nature, synchronization is not a concern.
We plan to use CSL asynchronous send/receive operations to implement migration.

Genomes will be represented as contiguous bufers of u32 and each PE's population as a contiguous buffer of genome data using a simple synchronous generational model (i.e., a swap buffer.)
Population-level operations (e.g., selection, reproduction, mutation) will take advantage of the CSL's data structure descriptors.
The local memory is sufficient for a population size on the order of hundreds or low thousands.
This could give a globl population size on the order of hundreds of millions.
